Speed and strength of an epidemic intervention

### 2020-03-02 doc/speedstrength.tex
Add the emailed comment
Ready for medR?

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}

@@

%% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.

@@

Next, we consider a ``test-and-treat'' strategy in which infected individuals are identified, linked to care and receive antiretroviral therapy (ART) with the goal of both preserving health and preventing transmission through viral suppression. \citep{garnett2009treating, granich2009universal, nah2017test}.
Our assumptions for this scenario are shown in \figref{test}.
We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because infected individuals don't have a way to know that they have HIV when they're infected) but increases very quickly (because sexually active individuals are the most likely to seek testing); 
after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment \figref{test}(C). The corresponding strength of intervention $L$ is shown in \figref{test}(A) and details of the assumption are given in the caption.

In this example, we see that the

The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important (\figref{test}D).
This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.

However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
(\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength (\figref{test}B).
In this scenario, we predict that the intervention remains effective over the range of considered parameters.

Using the speed-based decomposition provides a valuable perspective on this situation.
the assumed hazard rate goes down after a few months to reflect the saturation assumptions discussed above (\figref{test}C). 
The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
The effective speed of the intervention stays relatively constant, in part because the backward generation-interval distributions for different  scenarios are relatively similar.
The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important (\figref{test}D).
This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.

@@

\swp{can we combine the first two paragraphs into one paragraph? The second paragraph is just 2 sentences long...}
The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \RR, or outbreak ``strength'' -- below 1. The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \RR\, or even overlooked entirely \citep{park2020reconciling}.
We argue that \RR\ and $r$ provide equally valid, complementary perspectives on epidemic control, and that there are situations where each provides a clearer picture than the other.

@@

While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of the outcome than the other.
For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling},
Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universa}, or contact-tracing and quarantine for COVID-19; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level.
\swp{I switched the orders between speed and strength to remain consistent with the way previous sentences are organized.}
In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.

>>>>

\newcommand{\comment}{}
\renewcommand{\comment}[3]{}
\renewcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}

@@

\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}

@@

Next, we consider a ``test-and-treat'' strategy in which infected individuals are identified, linked to care and receive antiretroviral therapy (ART) with the goal of both preserving health and preventing transmission through viral suppression. \citep{garnett2009treating, granich2009universal, nah2017test}.
Our assumptions for this scenario are shown in \figref{test}.
We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because there is no way for newly infected individuals to know that they have HIV) but increases very quickly (because sexually active individuals are the most likely to seek testing); 
after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment \figref{test}(C). The corresponding strength of intervention $L$ is shown in \figref{test}(A) and details of the assumption are given in the caption.

In this example, we see that, as \pEarly\ goes down and our estimate of epidemic strength increases, the estimate of intervention strength increases roughly in parallel. The increase in intervention strength makes sense, as we have more time to reach people on average before they transmit: this is the core of the result of \cite{eaton2014proportion}. 
In our scenario, we predict that the intervention remains effective over the range of considered parameters.

\jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}

Though there is a clear intuition for why both strengths increase as early transmission goes down, the speed paradigm provides insight into why these two increases are so close to parallel.
The estimated epidemic speed depends only on the observed growth rate -- it does not change if we change our assumption about the proportion of early transmission.
For the test-and-treat intervention, the effective epidemic speed also stays relatively constant (\figref{test}D), in part because we have (plausibly) assumed that hazard stays relatively constant for a few key months, and in part because the backward generation-interval distributions for different  scenarios are relatively similar.
The intervention speed increases slightly as proportion of early transmission increases because the subpopulation that the intervention fails to reach become relatively more important if late transmission is more important.
Thus, the speed paradigm provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission.

However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
(\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength (\figref{test}B).

@@

The effectiveness of an epidemic intervention is often measured by its ability to reduce the reproductive number -- \RR, or outbreak ``strength'' -- below 1. The exponential growth rate -- $r$, or outbreak ``speed'' -- is often seen just as a stepping stone to \RR\, or even overlooked entirely \citep{park2020reconciling}.
We argue that \RR\ and $r$ provide equally valid, complementary perspectives on epidemic control, and that there are situations where each provides a clearer picture than the other.

@@

While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of the outcome than the other.
For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling},
Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universa}, or contact-tracing and quarantine for COVID-19; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level, like condom use.
In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.
